---
  output: github_document
  title: "ctscore: Contact Tracing Scoring System"
---

  <!-- README.md is generated from README.Rmd. Please edit that file. -->
  <!-- The code to render this README is stored in .github/workflows/render-readme.yaml -->
  <!-- Variables marked with double curly braces will be transformed beforehand: -->
  <!-- `packagename` is extracted from the DESCRIPTION file -->
  <!-- `gh_repo` is extracted via a special environment variable in GitHub Actions -->


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%", 
  dpi = 300
)
```


## Getting started

To install the package from github:
```{r eval = FALSE}
pak::pkg_install("thibautjombart/ctscore")
```


## Worked example

In the following, we read a toy dataset included in the package, which contains simulated contact tracing data for 30 contacts. Each contact has one or more exposures, some of which led to infection. Data are stored inside an xlsx file distributed with the package, but in practice you would read your own data from a csv or xlsx file.

```{r}
library(ctscore)
library(rio)
library(magrittr)
library(tibble)
library(dplyr)
library(ggplot2)

## use the path to your own file in practice 
path_to_file <- system.file("toy_ctdata.xlsx", package = "ctscore")

## read the data in
raw_data <- rio::import(path_to_file) %>% 
  tibble()

raw_data

```

Each row corresponds to a reported exposure. The file contains the following information:

- `contact_id`: a unique identifier for each contact
- `date`: the date of exposure, indicated as an integer since the first exposure; in practice, this could be actual dates in the format YYYY-MM-DD
- `type`: the type of exposure; here, either "household" or "funeral"
- `location`: the location of the contact; here, either "new_city", "local_town", or "hotspot"; these would be actual names of places in practice
- `last_visit`: the date of the last visit to the contact, where they exhibited no symptoms; if the contact has not been visited yet, this is `NA`
- `infected`: the simulation truth, indicating whether the contact was infected or not; this is not known in practice, but is included here for demonstration purposes
- `onset`: the simulation truth, indicating the day of symptom onset for infected contacts; this may is not known in practice, but is included here for demonstration purposes

Now that we have imported the data into R, we need to convert them to a `ctdata` object using `make_ctdata()` (see ?make_ctdata for details). 

Infection probabilities (argument `infection_proba`) can be estimated from previous contact tracing data, from the literature, or by expert opinion. Here, we assume that household exposures have a 20% probability of infection, while funeral exposures have a 90% probability of infection.

```{r}
x <- make_ctdata(
  contact_id = raw_data$contact_id,
  date = raw_data$date,
  type = raw_data$type,
  location = raw_data$location,
  infection_proba = list(household = 0.2, funeral = 0.9),
  last_visit = raw_data$last_visit
)

head(x)
class(x)
```

Smaller datasets can be visualised using `plot()`:
```{r}
plot(x)
```


Next, we need to specify incubation time distribution, which can be provided either as a vector of probabilities (giving p(0 day), p(1 day), p(2 days) ...) or as a `distcrete` object as returned by distcrete::distcrete(). Here, we generate a dummy incubation time distribution as a discretized Gamma:

```{r}
incub <- distcrete::distcrete(
  "gamma", 
  interval = 1, 
  shape = 3.5, 
  scale = 2.5, w = 0
)
plot(
  incub$d(0:30), type = "h", lwd = 6, lend = 1, 
  xlab = "Days since infection", 
  ylab = "Probability of symptom onset",
  main = "Incubation time distribution",
  col = 2
)

```

We can now calculate the `ctscore` using the `ctscore()` function, the current date is day 31 (again, this could be a real date in practice, in YYYY-MM-DD format):

```{r}
score <- ctscore(x, incub, t = 31)
score
```

`score` indicates, for each contact, the probability that a visit today will lead to detecting a new case. For convenience, we can append this score to the contact tracing database, and reorder contacts by decreasing order of priority/score:

```{r}
## adding scores to the original exposures table
res_full <- left_join(x, 
                      tibble(contact_id = as.integer(names(score)),
                             score = score), 
                      by = "contact_id") %>% 
  arrange(desc(score))
res_full

## same but keeping only contact information (no exposures), simpler to
## handle to prioritise individuals
res <- left_join(
  tibble(contact_id = unique(res_full$contact_id)), 
  tibble(contact_id = as.integer(names(score)), score = score)
  ) %>% 
    arrange(desc(score))
res


## some wrangling needed to keep the order of contacts in the plot
res %>% 
  mutate(contact_id = factor(contact_id, levels = unique(contact_id))) %>% 
ggplot(aes(x = score, y = contact_id)) + 
  geom_col() + 
  theme_bw() + 
  labs(x = "ctscore (probability of detecting symptoms)", 
       y = "Contact ID", 
       title = "Contact tracing scoring")

```


