---
  output: github_document
  title: "ctscore: Contact Tracing Scoring System"
---

  <!-- README.md is generated from README.Rmd. Please edit that file. -->
  <!-- The code to render this README is stored in .github/workflows/render-readme.yaml -->
  <!-- Variables marked with double curly braces will be transformed beforehand: -->
  <!-- `packagename` is extracted from the DESCRIPTION file -->
  <!-- `gh_repo` is extracted via a special environment variable in GitHub Actions -->


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 300
)
```


## Getting started

To install the package from github:
```{r eval = FALSE}
pak::pkg_install("thibautjombart/ctscore")
```


## Worked example

### Raw data

In the following, we read a toy dataset included in the package, which contains simulated contact tracing data for 30 contacts. Each contact has one or more exposures, some of which led to infection. Data are stored inside xlsx files distributed with the package, but in practice you would read your own data from csv or xlsx files.

```{r}
library(ctscore)
library(rio)
library(tibble)
library(dplyr)
library(ggplot2)

## use the path to your own file in practice

linelist <- rio::import(system.file(
  "toy_linelist.xlsx",
  package = "ctscore"
)) |>
  tibble()

linelist

exposures <- rio::import(
  system.file("toy_exposures.xlsx", package = "ctscore")
) |>
  tibble()

exposures
```


### Creating a `ctdata` object

A `ctdata` object is a list containing two data frames:

- `linelist`: A data frame containing individual-level information, one row per `contact_id`. It contains the following columns:
  - `contact_id`: a unique identifier for each contact (required)
  - `location`: the location of the contact (optional)
  - `last_visit`: the date of the last visit to the contact (optional)
  - `infected`: a logical indicating whether the contact was infected (optional)
  - `onset_date`: the date of symptom onset (optional)
  
- `exposures`: A data frame containing exposure-level information, one row per exposure. It contains the following columns:
  - `contact_id`: a unique identifier for each contact (required)
  - `date`: the date of exposure, indicated as an integer since the first exposure; in practice, this could be actual dates in the format YYYY-MM-DD (required)
  - `type`: the type of exposure (e.g. "household", "funeral") (required)
  - `infection_proba`: the probability of infection for a type of exposure (derived from the `infection_proba` argument)


Infection probabilities (argument `infection_proba`) can be estimated from previous contact tracing data, from the literature, or by expert opinion. Here, we assume that household exposures have a 20% probability of infection, while funeral exposures have a 90% probability of infection.

```{r}
x <- make_ctdata(
  linelist = linelist,
  exposures = exposures,
  infection_proba = list(household = 0.2, funeral = 0.9)
)
x
class(x)
```

Smaller datasets can be visualised using `plot()`:
```{r}
plot(x)
```

### Scoring with `ctscore()`

`ctscore` computes, for each contact, the probability that a visit today will lead to detecting a new case. 

First, we need to specify incubation period distribution, which can be provided either as a vector of probabilities (giving p(0 day), p(1 day), p(2 days) ...) or as a `distcrete` object as returned by distcrete::distcrete(). Here, we generate a dummy incubation time distribution as a discretized Gamma:

```{r}
incub <- distcrete::distcrete(
  "gamma",
  interval = 1,
  shape = 3.5,
  scale = 2.5,
  w = 0
)

plot(
  incub$d(0:30),
  type = "h",
  lwd = 6,
  lend = 1,
  xlab = "Days since infection",
  ylab = "Probability of symptom onset",
  main = "Incubation time distribution",
  col = 2
)
```

We can now calculate the `score` using the `ctscore()` function, the current date is day 31 (again, this could be a real date in practice, in YYYY-MM-DD format):

```{r}
score <- ctscore(x, incub, current_date = 31)
score
```


For prioritising individuals, `add_ctscore()` attaches the scores onto the `linelist` of the `ctdata` object:

```{r}
## ctdata with a `score` column in the linelist
xs <- add_ctscore(x, score)
```

We can use a simple plot to visualise the scores for each contact, which can be useful for prioritising visits:

```{r}
xs$linelist |>
  ggplot(aes(x = score, y = reorder(contact_id, score))) +
  geom_col() +
  theme_bw() +
  labs(
    x = "ctscore (probability of detecting symptoms)",
    y = "Contact ID",
    title = "Contact tracing scoring"
  )
```


We can use `as_tibble()` to flatten the `ctdata` object into a single data frame.
The argument `by_contact` indicates whether to return one row per contact (TRUE) or one row per exposure (FALSE, default):

```{r}
as_tibble(xs, by_contact = TRUE) |>
  slice_max(score, n = 1)

as_tibble(xs) |>
  filter(contact_id == 13)
```

The contact (ID 13) has the highest score because they had multiple 
exposures, including several funeral exposures, and were last seen 26 days ago (on day 5, current day is 31), so there was ample time for symptoms to have appeared since then. In fact, the linelist recorded that this contact was infected and developed symptoms on day 14.

```{r}
as_tibble(xs, by_contact = TRUE) |>
  slice_min(score, n = 1) |>
  _[["exposures"]]
```

The lowest scoring contact had a single 'weak' 
exposure only 2 days ago, so that it is unlikely that the individual has been infected, and if they have been, it is unlikely that they have developed symptoms yet. 


## Simulations

## Simulating contact tracing data

We can simulate contact tracing data using the `sim_ctdata()` function. This function generates a synthetic dataset of contacts and exposures, which can be useful for testing different follow-up strategies and evaluating the performance of the scoring system.
See `?sim_ctdata` for details on the arguments.

```{r}
set.seed(123)
sim <- sim_ctdata(
  n_contacts = 20,
  duration = 30,
  incub = incub$r(100),
  locations = list(cityA = 0.7, cityB = 0.3),
  n_exposures = list(cityA = 2, cityB = c(2, 2, 3, 4, 5, 10)),
  infection_proba = list(household = 0.2, funeral = 0.4),
  type_proba = list(household = 0.7, funeral = 0.3)
)
sim
```

`last_visit_date` is empty as we have not yet simulated a follow-up strategy. We can simulate a follow-up strategy using the `sim_followup()` function.

```{r}
f <- sim |>
  sim_followup(
    time = 30,
    delay = 5,
    duration = incub$q(0.99),
    coverage = 0.2,
    strategy = "random"
  )
f |> as_tibble(by_contact = TRUE)
```


```{r}
plot(f) +
  facet_wrap(~location, scales = "free_y")
```