---
  output: github_document
  title: "ctscore: Contact Tracing Scoring System"
  author: Thibaut Jombart
---

  <!-- README.md is generated from README.Rmd. Please edit that file. -->
  <!-- The code to render this README is stored in .github/workflows/render-readme.yaml -->
  <!-- Variables marked with double curly braces will be transformed beforehand: -->
  <!-- `packagename` is extracted from the DESCRIPTION file -->
  <!-- `gh_repo` is extracted via a special environment variable in GitHub Actions -->


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%", 
  dpi = 300
)
```


## Getting started

To install the package from github:
```{r eval = FALSE}
pak::pkg_install("thibautjombart/ctscore")
```


## Worked example

In the following, we read a toy dataset included in the package, which contains simulated contact tracing data for 30 contacts. Each contact has one or more exposures, some of which led to infection. Data are stored inside an xlsx file distributed with the package, but in practice you would read your own data from a csv or xlsx file.

```{r}
library(ctscore)
library(rio)
library(magrittr)
library(tibble)
library(dplyr)
library(ggplot2)

## use the path to your own file in practice 
path_to_file <- system.file("toy_ctdata.xlsx", package = "ctscore")

## read the data in
raw_data <- rio::import(path_to_file) %>% 
  tibble()

raw_data

```

Each row corresponds to a reported exposure. The file contains the following information:

- `contact_id`: a unique identifier for each contact
- `date`: the date of exposure, indicated as an integer since the first exposure; in practice, this could be actual dates in the format YYYY-MM-DD
- `type`: the type of exposure; here, either "household" or "funeral"
- `location`: the location of the contact; here, either "new_city", "local_town", or "hotspot"; these would be actual names of places in practice
- `last_visit`: the date of the last visit to the contact, where they exhibited no symptoms; if the contact has not been visited yet, this is `NA`
- `infected`: the simulation truth, indicating whether the contact was infected or not; this is not known in practice, but is included here for demonstration purposes
- `onset`: the simulation truth, indicating the day of symptom onset for infected contacts; this may is not known in practice, but is included here for demonstration purposes

Now that we have imported the data into R, we need to convert them to a `ctdata` object using `make_ctdata()` (see ?make_ctdata for details). 

Infection probabilities (argument `infection_proba`) can be estimated from previous contact tracing data, from the literature, or by expert opinion. Here, we assume that household exposures have a 20% probability of infection, while funeral exposures have a 90% probability of infection.

```{r}
x <- make_ctdata(
  contact_id = raw_data$contact_id,
  date = raw_data$date,
  type = raw_data$type,
  location = raw_data$location,
  infection_proba = list(household = 0.2, funeral = 0.9),
  last_visit = raw_data$last_visit
)

head(x)
class(x)
```

Smaller datasets can be visualised using `plot()`:
```{r}
plot(x)
```


Next, we need to specify incubation time distribution, which can be provided either as a vector of probabilities (giving p(0 day), p(1 day), p(2 days) ...) or as a `distcrete` object as returned by distcrete::distcrete(). Here, we generate a dummy incubation time distribution as a discretized Gamma:

```{r}
incub <- distcrete::distcrete(
  "gamma", 
  interval = 1, 
  shape = 3.5, 
  scale = 2.5, w = 0
)
plot(
  incub$d(0:30), type = "h", lwd = 6, lend = 1, 
  xlab = "Days since infection", 
  ylab = "Probability of symptom onset",
  main = "Incubation time distribution",
  col = 2
)

```

We can now calculate the `ctscore` using the `ctscore()` function, the current date is day 31 (again, this could be a real date in practice, in YYYY-MM-DD format):

```{r}
score <- ctscore(x, incub, current_date = 31)
score
```

`score` indicates, for each contact, the probability that a visit today will
lead to detecting a new case. 

For convenience, we can ask `ctscore ` to return results in other formats through the argument `out_type`:

- "ctdata": returns a `ctdata` object with only individual data and scores 
(stripping exposures); this is the most useful for prioritising individuals for 
visits

- "data.frame": returns a `data.frame` with contact IDs and scores

- "ctdata_full": a `ctdata` object with an additional column for the scores 
appended to the exposure database

```{r}
## `ctdata` with individual info and  scores, sorted by score
res <- ctscore(x, incub, current_date = 31, out_type = "ctdata") %>% 
    arrange(desc(score))
head(res)

## ctdata object with scores appended, sorted by score
res_full <- ctscore(x, incub, current_date = 31, out_type = "ctdata_full") %>% 
  arrange(desc(score))

```

We can use a simple plot to visualise the scores for each contact, which can be useful for prioritising visits:

```{r}

## some wrangling needed to keep the order of contacts in the plot
res %>% 
  mutate(contact_id = factor(contact_id, levels = unique(contact_id))) %>% 
ggplot(aes(x = score, y = contact_id)) + 
  geom_col() + 
  theme_bw() + 
  labs(x = "ctscore (probability of detecting symptoms)", 
       y = "Contact ID", 
       title = "Contact tracing scoring")
```


We can look at the first and last scores to understand how the scoring system works. 

```{r}
head(res_full)
```

The first contact (ID 14) has the highest score because they had multiple 
exposures, including several funeral exposures, and were last seen 16 days ago 
(on day 15, current day is 31), so there was ample time for symptoms to have appeared since then. The second contact (ID 13) has a high score for the same reasons: multiple exposures, and a last visit a long time ago.


```{r}
tail(res_full)
```

The very last contact (ID 20) has a low score because of a single 'weak' 
exposure only 2 days ago, so that it is unlikely that the individual has been infected, and if they have been, it is unlikely that they have developed 
symptoms yet. 

The second last contact (ID 29) reported a high-risk (funeral) exposure, but 
this was very recent (day 29) so they are unlikely to show symptoms just yet.

The situation is the same for the third last contact (ID 21), whose funeral 
exposure is yet too recent to show symptoms just yet, if they were infected. 

The fourth last contact (ID 19) has reported multiple weak exposures, but they 
were last visited yesterday and showed no symptoms, so they are unlikely to be infected.
